\section{Appendix A : Metadata Structures}
\label{sec:appendixA}

The \texttt{AssetMetaData} data type contains the following fields outlined below:

\begin{itemize}
 
\item \texttt{assetModuleType}: contains the module name where are the definitions of abstract data types of the \textit{<newAsset>}. Format: \texttt{<NewAssetDirectory>.Types}

\item \texttt{assetModuleParser}: contains the module name where is implemented the \textit{parser} function of the \texttt{<newAsset>}. Format: \texttt{<NewAssetDirectory>.Parsers.<NewAssetFomartParser>}

\item \texttt{assetModuleType'}: contains the module name where are the definitions of \texttt{SPLModel}, \texttt{InstanceModel}, \texttt{TransformationModel} and \texttt{ConfigurationKnowledge} data types. This module must be imported into generated Hephaestus instance module. Format: \texttt{TestTypes}

\item \texttt{assetModule}: is the module name containing the definitions the \texttt{empty<NewAsset>} and \texttt{transform<NewAsset>} functions. This module must be located below \texttt{HplAssets} directory. Format: \texttt{<NewAsset>}

\item \texttt{assetModel}: is the data type which defines the \texttt{<newAsset>} located in \texttt{Types.hs} module in the \texttt{HplAssets/<NewAssetDirectory>} directory. This data type is used to extend \texttt{SPLModel} and \texttt{InstanceModel} data types in Hephaestus empty instance. Format: \texttt{<NewAsset>Model}

\item \texttt{assetSelector}: is a list of tuples containing the field's name and type to extend the \texttt{InstanceModel} data type to \textit{<newAsset>}. Format: \texttt{[(<nameField>,<typeField>),...]}

\item \texttt{assetSelector'}: is a list of tuples containing the field's name and type to extend the \texttt{SPLModel} data type to \textit{<newAsset>}. Format: \texttt{[(<nameField>,<typeField>),...]}

\item \texttt{assetEmpty}: function name which defines the empty product of the \texttt{<NewAsset>}. Format: \texttt{empty<newAsset>}

\item \texttt{assetXType}: data type's name which defines the constructors of the \texttt{<NewAsset>} transformations. Format: \texttt{<NewAsset>Transformation}

\item \texttt{assetXFun}: function name which defines the implementation of the \texttt{<NewAsset>} transformation functions, located in the \texttt{<newAsset>} module in \texttt{HplAssets} directory. Format: \texttt{transform<newAsset>}

\item \texttt{assetVarProperty}: variable name used to introduce the \texttt{findPropertyValue} instruction into \texttt{main} function of Hephaestus instance to reading the file name which contains the artifacts of \texttt{<NewAsset>} product line. Format: \texttt{<anyString>}

\item \texttt{assetNameProperty}:









\end{itemize}