\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

\newcommand{\hpl}{\textsf{Hephaestus-PL}}
\newcommand{\hp}{\textsf{Hephaestus}}
\newcommand{\hpbase}{\hp{} \emph{Base}}
\newcommand{\hpsplasset}{\hp{} \emph{SPL Asset}}
\newcommand{\hpproduct}{\hp{} \emph{Product}}

\usepackage{amssymb}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{subfigure}
\usepackage{listings}
\usepackage{listings}
\usepackage{url}

\definecolor{gray}{RGB}{220,220,220}
\journal{Science of Computer Programming}

%\lstloadlanguages{Haskell}
%\lstnewenvironment{code}
%    {\lstset{}%
%      \csname lst@SetFirstLabel\endcsname}
%    {\csname lst@SaveFirstLabel\endcsname}
%    \lstset{
%      basicstyle=\small\ttfamily,
%      backgroundcolor=\color{gray},
%      flexiblecolumns=false,
%      basewidth={0.5em,0.45em},
%      literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
%               {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
%               {\\\\}{{\char`\\\char`\\}}1
%               {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
%               {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
%               {\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
%               {>>}{{>>}}2 {>>=}{{>>=}}2
%               {|}{{$\mid$}}1
%    }

\lstdefinelanguage{haskell}{%
%  numbers=none,
  morekeywords={do,module,qualified,import,let,case,of,in,class,instance,data,newtype,where,otherwise,deriving,type,extends,public,forall},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
}

\lstset{%
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it,
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\footnotesize\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}


\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Configurability and Extensibility of Tools for\\ Software Product
Lines}
%\title{\hpl: A Product Line for Tools for Product Derivation}
%\title{A configurable, flexible, and extensible tool for software-product derivation}
%\title{Supporting Configurability and Flexibility in Software Product Line Tool Development: the \hpl{} Case}
%\title{On the Design and Implementation of \hpl: a Software Product Line of Software Product Line Tools}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[unb]{Lucin\'{e}ia Turnes}
\ead{lucineiaturnes@gmail.com}
\author[unb]{Vander Alves\corref{cor1}}
\author[unb]{Rodrigo Bonif\'{a}cio}
\ead{rbonifacio@cic.unb.br}
\cortext[cor1]{Corresponding author. Telephone: +55 61 8157-1464. Fax: +55 61 3107-7414.}
\ead{valves@unb.br}
\author[ukob]{Ralf L\"{a}mmel}
\ead{rlaemmel@acm.org}

\address[unb]{Departamento de Ci\^{e}ncia da Computa\c{c}\~{a}o, Universidade de Bras\'{i}lia, Campus Darcy Ribeiro, Bras\'{i}lia, Brazil, 70910-900}
\address[ukob]{Software Languages Team, Universit\"{a}t Koblenz-Landau, Institut f\"{u}r Informatik, Universittsstra\ss{}e 1, D-56070 Koblenz, Germany}

\author{}

\address{}

\begin{abstract}

% Context
  Tool support is essential for application engineering in software
  product lines for the derivation of products from reusable assets.
% Problem
  Existing tools are insufficiently configurable and extensible with
  regard to the type of artifacts and associated variability mechanisms
  that they target.
% Solution
  This paper addresses the problem by exploring the underlying
  commonality and adequately managing variability of tools for product
  derivation. More specifically, the paper describes domain analysis,
  design, implementation of \hpl---a proper product line for such
  tools. To this end, an extractive strategy was applied to the
  existing Haskell-based \hp{} tool. A transformational approach to
  variability management is designed and implemented by
  metaprogramming in Haskell. We also describe a reactive process for
  evolving \hpl{} to address new types of artifacts.
% Evidence
  Improved configurability and extensibility is substantiated
  by an assessment.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
  Software Product Line Tools \sep Variability Management \sep \hp
  \sep Configurability \sep Extensibility \sep Bootstrapping \sep
  Expression Problem \sep Metaprogramming \sep Haskell

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

\input {introduction}
\input {hephaestus}
\input {domainAnalysis}
\input {domainDesign}
\input {implementation}
\input {process}
\input {resultDiscussion}
\input {relatedWork}
\input {conclusion}

\bibliographystyle{elsarticle-num}
\bibliography{references}

\end{document}
